The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-31 pkg/partykit/inst/guideliketest/draft/draft.Rnw
further revision based on feedback
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2603 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 9618f197678a5f0698391a05193202ce650cc61a

\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
To investigate the impact of each of the building blocks separately the most general 
case of the ``stump'' scenario where the intercept and slope parameter are both varying is 
considered. In the underlying evaluation all possible combinations of the building blocks 
have been included. The different levels of each of the three building blocks are listed in 
Table~\ref{tab:levels} where $h$ and $g$ refer to the transformation functions applied to the 
response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
In the case of categorization, the split variable $Z$ is binned at the
quartiles. This corresponds to a four-dimensional 0/1 transformation
function $g(Z)$ that indicates into which of the bins each observation falls.
Maximum selection across potential split points in a variable $Z$ also
corresponds to a multivariate 0/1 transformation function $g(Z)$. However,
in this case for each potential split point an indicator is used that is 0
before and 1 after the respective split point.
(See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
for a more detailed overview of all 12 combinations.)

@@

Both of the transformation functions $h$ and $g$ (applied to the response~$Y$ or a split 
variable~$Z$ respectively) can also be multivariate. For this case, the corresponding
statistical tests are available, in particular the \textsf{R} package \textbf{coin}
which is employed for the implementation of MOB and CTree provides 
the required toolbox for multivariate dependency tests, including a $\chi^2$~test.

@@

Also in more general models residual-based and score-based procedures are expected to perform
equally well if all parameters are highly correlated. On the other hand, if parameters are 
orthogonal changes that are not connected to the mean might be missed when only considering
residuals. This can be prevented by including scores which are a standard measurement of goodness 
of fit that can be easily employed for all model types. Moreover, the score function $s$ can 
simply be seen as a  transformation to a different space in order to allow for a well structured
analysis of dependencies, regardless of the type of model and therefore applicable in a
wide variety of situations.
In order to benefit from the advantages of using scores 
some already existing models have been adapted, e.g., in \cite{Hothorn+Zeileis:2017} 
and \cite{Schlosser+Hothorn+Stauffer:2019}. Similarly, variations of the GUIDE algorithm
using different types of residuals are provided, e.g., in \citealp{Loh+Zheng:2013}, 
\citealp{Chaudhuri+Loh:2002} and \citealp{Loh+He+Man:2015}.
But even though it also depends on the type of the applied residuals whether parameter changes 
can be detected successfully, a one-dimensional measurement of goodness of fit such as residuals
provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of testing all parameters.

>>>>

\subsection{``Stump'' scenario: Full factorial analysis of building blocks}\label{sec:3way}
To investigate the impact of each of the building blocks separately the most general 
case of the ``stump'' scenario where the intercept and slope parameter are both varying is 
considered. In the underlying evaluation all possible combinations of the building blocks 
have been included. The different levels of each of the three building blocks are listed in 
Table~\ref{tab:levels} where $h$ and $g$ refer to the transformation functions applied to the 
response $Y$ or a split variable $Z$ respectively, both as described in Section~\ref{sec:ctree}. 
In the case of categorization, the split variable $Z$ is binned at the
quartiles. This corresponds to a four-dimensional 0/1 transformation
function $g(Z)$ that indicates into which of the bins each observation falls.
Maximum selection across potential split points in a variable $Z$ also
corresponds to a multivariate 0/1 transformation function $g(Z)$. However,
in this case for each potential split point an indicator is used that is 0
before and 1 after the respective split point.
(See Table~\ref{tab:combinations_appendix} in Appendix~\ref{app:combinations} 
for a more detailed overview of all 12 combinations of the building blocks.)

@@

Both of the transformation functions $h$ and $g$ (applied to the response~$Y$ or a split 
variable~$Z$ respectively) can also be multivariate. For this case, the corresponding
statistical tests are available, in particular in the \textsf{R} package \textbf{coin}
which is employed for the implementation of MOB and CTree and provides 
the required toolbox for multivariate dependency tests, including a $\chi^2$~test.

@@

Also in more general models residual-based and score-based procedures are expected to perform
equally well if all parameters are highly correlated. On the other hand, if parameters are 
orthogonal changes that are not connected to the mean might be missed when only considering
residuals. This can be prevented by including scores which are a standard measurement of goodness 
of fit that can be easily employed for all model types. Moreover, the score function~$s$ can 
simply be seen as a transformation to a different space in order to allow for a well structured
analysis of dependencies, regardless of the type of model and therefore applicable in a
wide variety of situations.
In order to benefit from the advantages of using scores 
some already existing models have been adapted, e.g., in \cite{Hothorn+Zeileis:2017} 
and \cite{Schlosser+Hothorn+Stauffer:2019}. Similarly, variations of the GUIDE algorithm
using different types of residuals are provided, e.g., in \citealp{Loh+Zheng:2013}, 
\citealp{Chaudhuri+Loh:2002} and \citealp{Loh+He+Man:2015}.
But even though it also depends on the type of the applied residuals whether parameter changes 
can be detected successfully, a one-dimensional measurement of goodness of fit such as residuals
provides less power than full model scores. 
Whether this difference affects the performance of a testing strategy and if so, to what extent, is 
determined by the way how residuals are connected to parameter shifts as well as the type of parameter
shift and the type of model. Therefore, a general statement for all types of models and residuals can 
hardly be made. However, the results from previous literature such as \cite{Ploberger+Kraemer:1992} and
the here presented simulation show that considering only residuals provides small gains in certain 
situations but remarkable deficits in general. On the contrary, the benefits of including full model 
scores clearly outweigh the price of testing all parameters.
